% To be completed with Google Sheet

FSap would like to thank Jonathan Taylor, Ryan Giordano, Alexander Strang, and Olivier Bonte for useful comments and feedback. 
FSap and FP benefits from the Jupyter meets the Earth project supported by the NSF Earth Cube Program under awards 1928406, 1928374. 
JB has been supported by the Nederlandse Organisatie voor Wetenschappelijk Onderzoek, Stichting voor de Technische Wetenschappen (Vidi grant 016.Vidi.171.063).
BG acknowledges the support of the Helmholtz Einstein International Berlin Research School in Data Science as well as the Center for Scalable Data Analytics and Artificial Intelligence Dresden/Leipzig.
PH was supported in part by NSF CSSI \#OAC-2103942, ONR \#N00014-20-1-2772, DOE \#DE-SC002317, and a JPL/Caltech subcontract of the NASA Estimating the Circulation and Climate of the Ocean (ECCO) project.
GH was supported by the NSF grant DEB-1933497. 

Fsch, AP, and CR works was supported in part by the Director, Office of Science, Office of Advanced Scientific Computing Research, U.S. Department of Energy under Contract No. DE-AC02-05CH11231, and in part by the National Science Foundation under Grant DMS-2309596.
This material is based upon work supported by the Department of Energy, National Nuclear Security Administration under Award Number DE-NA0003965.
This report was prepared as an account of work sponsored by an agency of the United States Government. 
Neither the United States Government nor any agency thereof, nor any of their employees, makes any warranty, express or implied, or assumes any legal liability or responsibility for the accuracy, completeness, or usefulness of any information, apparatus, product, or process disclosed, or represents that its use would not infringe privately owned rights. 
Reference herein to any specific commercial product, process, or service by trade name, trademark, manufacturer, or otherwise does not necessarily constitute or imply its endorsement, recommendation, or favoring by the United States Government or any agency thereof. 
The views and opinions of authors expressed herein do not necessarily state or reflect those of the United States Government or any agency thereof. 
This material is based upon work supported by the National Science Foundation under grant no. OAC-2103804 , no. OSI-2029670, no. DMS-2325184, no. PHY-2028125. 
This material was supported by The Research Council of Norway and Equinor ASA through the Research Council project ”308817 - Digital wells for optimal production and drainage”. 
Fsch, AP, and CR research was sponsored by the United States Air Force Research Laboratory and the United States Air Force Artificial Intelligence Accelerator and was accomplished under Cooperative Agreement Number FA8750-19-2-1000. 
The views and conclusions contained in this document are those of the authors and should not be interpreted as representing the official policies, either expressed or implied, of the United States Air Force or the U.S. Government. 
The U.S. Government is authorized to reproduce and distribute reprints for Government purposes notwithstanding any copyright notation herein. 
Fsch, AP, and CR  would like to thank DARPA for funding this work through the Automating Scientific Knowledge Extraction and Modeling (ASKEM) program, Agreement No. HR0011262087. The views, opinions and/or findings ex-pressed are those of the authors and should not be interpreted as representing the official views or policies of the Department of Defense or the U.S. Government.

% CR benefits work is supported by the Department of Energy, National Nuclear Security Administration under Award Number DE-NA0003965.
% CR work is supported by the National Science Foundation under grant no. OAC-2103804 , no. OSI-2029670, no. DMS-2325184, no. PHY-2028125. 
% This material was supported by The Research Council of Norway and Equinor ASA through the Research Council project "308817 - Digital wells for optimal production and drainage". 
% CR research was sponsored by the United States Air Force Research Laboratory and the United States Air Force Artificial Intelligence Accelerator and was accomplished under Cooperative Agreement Number FA8750-19-2-1000. 
% CR would like to thank DARPA for funding this work through the Automating Scientific Knowledge Extraction and Modeling (ASKEM) program, Agreement No. HR0011262087. 