Differential equations are mathematical objects used as scientific models to explicitly describe the processes and dynamics within various systems, based on prior domain knowledge. 
They are fundamental in many scientific disciplines for modeling phenomena such as physical processes, population dynamics, social interactions, and chemical reactions.
% \todo[inline]{here, it would be nice to explain how they differ to data-drive models. Consider:}
By contrast, data-driven models do not necessarily require a detailed understanding of the underlying physical processes, and learn patterns and relationships directly from data. 
By embedding assumptions on the system dynamics in their structure, traditional mathematical models predictions are constrained, requiring few data to be calibrated for forecasting a phenomenon. Yet, their structure makes calibration difficult, and they often fail at describing complex processes, rendering them less accurate.
On the other hand, data-driven models are particularly useful when the underlying processes are poorly understood or complex, but require a significantly larger amount of data.
The combination of mechanistic models with data-driven models is becoming increasingly common in many scientific domains. 
To achieve this, these models need to leverage both domain knowledge and data, to have an accurate representation of the underlying dynamics. 
Being able to determine which model parameters are most influential and further compute derivatives of such a model is key to correctly assimilating and learning from data, but a myriad of sensitivity methods exist to do so. 
In this review, we present an overview of the different sensitivity methods that exist, providing (i) guidelines on the best use cases for different scientific domain problems, (ii) detailed mathematical analyses of their characteristics, and (iii) computational implementations on how to solve them efficiently. 
