This is a list of the code provided along with the current manuscript.
All the following scripts can be found in the GitHub repository \href{https://github.com/ODINN-SciML/DiffEqSensitivity-Review}{\texttt{DiffEqSensitivity-Review}}. 
 
\begin{itemize}
    \item[$\clubsuit_\text{\ref{code:figure-comparison}}$] \textbf{Comparison of direct methods.} The script \href{https://github.com/ODINN-SciML/DiffEqSensitivity-Review/blob/main/code/DirectMethods/Comparison/direct-comparision.jl}{\texttt{direct-comparision.jl}} reproduces Figure \ref{fig:direct-methods}.
    \item[$\clubsuit_\text{\ref{code:dual-number}}$] \textbf{Dual numbers definition.} The script \href{https://github.com/ODINN-SciML/DiffEqSensitivity-Review/blob/main/code/DirectMethods/DualNumbers/dualnumber_definition.jl}{\texttt{dualnumber\_definition.jl}} includes a very simple example of how to define a dual number using \texttt{struct} in Julia and how to extend simple unary and binary operations to implement the chain rule usign multiple distpatch. 
    \item[$\clubsuit_\text{\ref{code:AD-wrong}}$] \textbf{When AD is algorithmically correct but numerically wrong.} The script \href{https://github.com/ODINN-SciML/DiffEqSensitivity-Review/blob/main/code/SensitivityForwardAD/example-AD-tolerances.jl}{\texttt{example-AD-tolerances.jl}} includes the example shown in Section \ref{section:software-Forward-AD} and further elaborated in Section \ref{section:AD-incorrect} where forward AD gives the wrong answer when tolerances in the gradient are not computed taking into account both numerical errors in the numerical solution and the sensitivity matrix. Further examples of this phenomena can be found in the  and the Julia \href{https://github.com/ODINN-SciML/DiffEqSensitivity-Review/blob/main/code/SensitivityForwardAD/testgradient_julia.jl}{\texttt{testgradient\_julia.jl}}.
    \item[$\clubsuit_\text{\ref{code:AD-wrong-JAX}}$] \textbf{When AD is algorithmically correct but numerically wrong (JAX)}. Python script \href{https://github.com/ODINN-SciML/DiffEqSensitivity-Review/blob/main/code/SensitivityForwardAD/testgradient_python.py}{\texttt{testgradient\_python.py}}
    \item[$\clubsuit_\text{\ref{code:complex-step}}$] \textbf{Complex step in numerical solver.} The script \href{https://github.com/ODINN-SciML/DiffEqSensitivity-Review/blob/main/code/DirectMethods/ComplexStep/complex_solver.jl}{\texttt{complex\_solver.jl}} shows how to define the dynamics of the ODE to support complex variables and then compute the complex step derivative. 
    \item[$\clubsuit_\text{\ref{code:sensitivity-equation}}$] \textbf{Solving the forward sensitivity equation. } The scrip \href{https://github.com/ODINN-SciML/DiffEqSensitivity-Review/blob/main/code/SolverMethods/Harmonic/forward_sensitivity_equations.jl}{\texttt{forward_sensitivity_equations.jl}} includes a manual implementation of the  forward sensitivity equations. This also includes how to compute the same sensitivity using \texttt{ForwardSensitivity} in Julia. 
    \item[$\clubsuit_\text{\ref{code:discrete-adjoint}}$] The script \href{https://github.com/ODINN-SciML/DiffEqSensitivity-Review/blob/main/code/SolverMethods/Harmonic/adjoint_discrete.jl}{\texttt{adjoint_discrete.jl}} includes a manual implementation of the discrete adjoint method for the simple harmonic oscillator. 
\end{itemize}